Learning complex models with invertible neural networks: a likelihood-free Bayesian approach

### 2019-06-28 latex/paper_main.tex
Fix SIR

For the following application, we set $d=4$, and $a=b=1$. The design matrices for each iteration contain a variable number $n$ of \textit{i.i.d.} data points drawn from a unit Gaussian $\mathcal{N}_{4}(\boldsymbol{0},\boldsymbol{I})$. The number of trials is drawn from a uniform distribution $n \sim \mathcal{U}(50, 500)$ at each training iteration (Lines $2$-$9$ of \textbf{Algorithm} \ref{alg:1}).

@@

\subsection*{Example 1 - The Ricker Model}
Discrete population dynamics models describe how the number of individuals in a population changes over discrete units of time \cite{wood2010statistical}. In particular, the Ricker model describes the number of individuals $x_{t+1}$ in generation $t+1$ as a function of the number of individuals in the previous generation $t$ by the following non-linear equation:
\begin{align*}
N_{t+1} &= rN_{t}\exp(-N_{t} + \epsilon_{t}) \numberthis \\
x_{t} &\sim Pois(\rho N_{t}) \numberthis
\end{align*}
for $t = 1,...,T$ where $N_{t}$ is the expected number of individuals at time $t$, $r$ is the growth rate, $\rho$ is a scaling parameter and $\epsilon_{t} \sim \mathcal{N}(0, \sigma^{2})$ is random Gaussian noise \cite{mestdagh2018prepaid}. The Ricker model has no known likelihood function and is thus a suitable candidate for likelihood-free inference. The parameter estimation task is thus to recover $\boldsymbol{\theta}=(\rho,r,\sigma)$ from the observed one-dimensional time-series data $\boldsymbol{x}=(x_{1},x_{2},...,x_{T})$ where each $x_{t} \in \mathbb{N}$.

During training of the networks, we simulate time-series from the Ricker model with varying lengths. The number of time points $T$ is drawn from a uniform distribution $T \sim \mathcal{U}(100, 500)$ at each training iteration (see the \textbf{SI} for more details about the simulation). What if the data does not contain information about a given parameter? In this case, any good estimation method should detect this, and return the prior of the particular parameter. To test this, we add a random uniform variable $u \sim \mathcal{U}(0, 1)$ to the parameter vector $\boldsymbol{\theta}$ and train the model with this additional dummy variable. Then, we inspect the posterior of $u$ for uniformity.

@@

The results on the LFM model are depicted in \autoref{fig:Fig.4}.

@@

These advantages notwithstanding, some limitations of the method deserve a mention.  Even though high-level deep learning libraries, such as \textit{TensorFlow} or \textit{Torch}, allow for rapid and relatively straightforward development of various neural network architectures, the implementational burden associated with the current method is still reasonably high. In order to ease the understanding and independent application of the method, we provide fully functioning code to reproduce and study all of the examples tackled in this paper (\href{https://github.com/stefanradev93/cINN}{https://github.com/stefanradev93/cINN}). Moreover, we are currently developing a general user-friendly software, which should abstract away most of the methodological complexities from the user. Another potential shortcoming of the method is the seemingly overwhelming number of hyperparameters that might require fine-tuning by the user for optimal performance on a given task. However, we observe that many of the default values of hyperparameters are sufficient to achieve good performance, and starting with a relatively large default network of $10$ ACBs does not appear to hurt performance or destabilize training. We expect that a single architecture should be able to perform well on almost all models from a given domain (i.e., a single architecture for decision-making models). Future research should investigate the question of generality by applying the method to challenging parameter estimation tasks across different research domains.

>>>>

For the following application, we set $d=4$, and $a=b=1$. The design matrices for each iteration contain a variable number $n$ of \textit{i.i.d.} data points drawn from a unit Gaussian $\boldsymbol{x}^{(i)} \sim \mathcal{N}_{4}(\boldsymbol{0},\boldsymbol{I})$ for $i=1,...,n$. The number of trials is drawn from a uniform distribution $n \sim \mathcal{U}(50, 500)$ at each training iteration (Lines $2$-$9$ of \textbf{Algorithm} \ref{alg:1}).

@@

\subsection*{Example 1 - The Ricker Model}
Discrete population dynamics models describe how the number of individuals in a population changes over discrete units of time \cite{wood2010statistical}. In particular, the Ricker model describes the number of individuals $x_{t+1}$ in generation $t+1$ as a function of the number of individuals in the previous generation $t$ by the following non-linear equation:
\begin{align*}
x_{t} &\sim Pois(\rho N_{t}) \numberthis \\
N_{t+1} &= rN_{t}e^{-N_{t} + \epsilon_{t}} \numberthis 
\end{align*}
for $t = 1,...,T$ where $N_{t}$ is the expected number of individuals at time $t$, $r$ is the growth rate, $\rho$ is a scaling parameter and $\epsilon_{t} \sim \mathcal{N}(0, \sigma^{2})$ is random Gaussian noise \cite{mestdagh2018prepaid}. The Ricker model has no known likelihood function and is thus a suitable candidate for likelihood-free inference. The parameter estimation task is thus to recover $\boldsymbol{\theta}=(\rho,r,\sigma)$ from the observed one-dimensional time-series data $\boldsymbol{x}=(x_{1},x_{2},...,x_{T})$ where each $x_{t} \in \mathbb{N}$.

During training of the networks, we simulate time-series from the Ricker model with varying lengths. The number of time points $T$ is drawn from a uniform distribution $T \sim \mathcal{U}(100, 500)$ at each training iteration (see the \textbf{SI} for more details about the simulation).

What if the data does not contain information about a given parameter? In this case, any good estimation method should detect this, and return the prior of the particular parameter. To test this, we add a random uniform variable $u \sim \mathcal{U}(0, 1)$ to the parameter vector $\boldsymbol{\theta}$ and train the model with this additional dummy variable. Then, we inspect the posterior of $u$ for uniformity.

@@

The results on the LFM model are depicted in \autoref{fig:Fig.4}. XXX

@@

These advantages notwithstanding, some limitations of the method deserve a mention.  Even though high-level deep learning libraries, such as \textit{TensorFlow} or \textit{Torch}, allow for rapid and relatively straightforward development of various neural network architectures, the implementational burden associated with the current method is still reasonably high. In order to ease the understanding and independent application of the method, we provide fully functioning code to reproduce and study all of the examples tackled in this paper (\href{https://github.com/stefanradev93/cINN}{https://github.com/stefanradev93/cINN}). Moreover, we are currently developing a general user-friendly software, which should abstract away most of the methodological complexities from the user. Another potential shortcoming of the method is the seemingly overwhelming number of hyperparameters that might require fine-tuning by the user for optimal performance on a given task. However, we observe that many of the default hyperparameter values are sufficient to achieve excellent performance, and starting with a relatively large default network of $10$ ACBs does not appear to hurt performance or destabilize training, even if the model to be learned is relatively simple. We expect that a single architecture should be able to perform well on almost all models from a given domain (i.e., a single architecture for decision-making models). Future research should investigate the question of generality by applying the method to challenging parameter estimation tasks across different research domains.

### 2019-06-28 latex/paper_supplementary.tex
Fix SIR

\usepackage{algorithm}
\usepackage[noend]{algpseudocode}
\graphicspath{{figures/}}
\DeclareMathOperator{\Tr}{Tr}

@@

\subsubsection*{Simulation-Based Calibration}
Simulation-based calibration is a recently proposed method for validating the accuracy of posterior samples generated by a Bayesian sampling method \cite{talts2018validating}. It is based on the so called \textit{self-consistency} of the Bayesian joint distribution. Given a sample from the prior distribution $\tilde{\theta} \sim p(\theta)$ and a sample from the data-generating process $\tilde{x} \sim p(x|\tilde{\theta})$, one can integrate $\tilde{\theta}$ and $\tilde{x}$ out of the Bayesian joint distribution to recover back the prior of $\theta$:
\begin{align*}
p(\theta) &= \int p(\theta,\tilde{\theta},\tilde{x})d\tilde{x}d\tilde{\theta} \numberthis  \label{eqn:6} \\
&= \int p(\theta,\tilde{x}|\tilde{\theta})p(\tilde{\theta})d\tilde{x}d\tilde{\theta} \numberthis \label{eqn:7} \\
&= \int p(\theta|\tilde{x})p(\tilde{x}|\tilde{\theta})p(\tilde{\theta})d\tilde{x}d\tilde{\theta} \numberthis \label{eqn:8}
\end{align*}
If the Bayesian sampling method produces samples from the exact posterior, the equality implied by Eq.\ref{eqn:8} should hold regardless of the particular form of the posterior. Thus, any violation of this equality indicates some error incurred by the sampling method. The authors of \cite{talts2018validating} propose the following algorithm relying on rank-statistics for visually detecting such violations:

\begin{algorithm}
\caption{Simulation-based calibration}\label{alg:1}
\begin{algorithmic}[1]
\State {Lol}

\end{algorithmic}
\end{algorithm}

>>>>

\usepackage{bbm}
\usepackage{algorithm}
\usepackage[noend]{algpseudocode}
\graphicspath{{figures/}}
\DeclareMathOperator{\Tr}{Tr}

@@

\subsubsection*{Simulation-Based Calibration}
Simulation-based calibration is a recently proposed method for validating the accuracy of posterior samples generated by a Bayesian sampling method \cite{talts2018validating}. It is based on the so called \textit{self-consistency} of the Bayesian joint distribution. Given a sample from the prior distribution $\tilde{\theta} \sim p(\theta)$ and a sample from the data-generating process $\tilde{x} \sim p(x|\tilde{\theta})$, one can integrate $\tilde{\theta}$ and $\tilde{x}$ out of the Bayesian joint distribution to recover back the prior of $\theta$:
\begin{align*}
p(\theta) &= \int p(\theta,\tilde{\theta},\tilde{x})d\tilde{x}d\tilde{\theta} \numberthis  \label{eqn:6} \\
&= \int p(\theta,\tilde{x}|\tilde{\theta})p(\tilde{\theta})d\tilde{x}d\tilde{\theta} \numberthis \label{eqn:7} \\
&= \int p(\theta|\tilde{x})p(\tilde{x}|\tilde{\theta})p(\tilde{\theta})d\tilde{x}d\tilde{\theta} \numberthis \label{eqn:8}
\end{align*}
If the Bayesian sampling method produces samples from the exact posterior, the equality implied by Eq.\ref{eqn:8} should hold regardless of the particular form of the posterior. Thus, any violation of this equality indicates some error incurred by the sampling method. The authors of \cite{talts2018validating} propose \textbf{Algorithm} \ref{alg:1} for visually detecting such violations:

\begin{algorithm}
\caption{Simulation-based calibration (SBC) for a single parameter $\theta$ }\label{alg:1}
\begin{algorithmic}[1]
\For{$i = 1,...,n$}
\State {Sample $\tilde{\theta}^{(i)} \sim p(\theta)$}
\State {Simulate a dataset $\boldsymbol{x}^{(i)} = q(\tilde{\theta}^{(i)})$}
\State {Draw posterior samples $\{\theta^{(l)}\}_{l=1}^{L} \sim p(\theta|\boldsymbol{x}^{(i)})$} 
\State {Compute rank statistic $r^{(i)} = \sum_{l=1}^{L}\mathbbm{1}_{[\theta^{(l)}<\tilde{\theta}^{(i)}]}$}
\State {Store $r^{(i)}$}
\EndFor
\State {\textbf{end for}}
\State {Analyze the histogram for uniformity}
\end{algorithmic}
\end{algorithm}

\textbf{Algorithm} \ref{alg:1} is justified, since Eq.\ref{eqn:8} implies that the rank statistic defined in line $5$ should be uniformly distributed. Hence, any deviations from uniformity indicate some error in the approximate posterior.
